### 2014-12-13 text/Chapter05.tex
Proofreading and editing (Chapter 5)

As already discussed, the aim of this thesis is to identify abstractions for context-aware
programming languages. We attempt to find abstractions that are general enough to capture a wide
range of useful programming language features, but specific enough to let us identify
interesting properties of the languages.

In Chapter~\ref{ch:applications}, we identified two notions of context. We generalized the 
class of flat calculi that capture whole-context properties in Chapter~\ref{ch:flat}. 
In this chapter, we turn our attention to \emph{structural} coeffect calculi that capture
per-variable properties.

The flat coeffect system captures interesting use-cases (implicit parameters, liveness and
data-flow), but provides relatively weak properties. We can define its categorical semantics,
but the equational theory proofs had numerous additional requirements. For this reason, it
is worthwhile to consider structural systems in a separate chapter. We will see that 
structural coeffects have a number of desirable properties that hold for all instances of the
calculus.

@@

Two examples of flat systems from the previous chapter were liveness and data-flow. As discussed
in \ref{sec:applications-structural}, these are interesting for theoretical reasons. However, tracking
liveness of the whole context is not practically useful. Structural versions of liveness and 
data-flow let us track more fine-grained properties. Moreover, the equational theory of flat 
coeffect calculus did not reveal many useful properties for flat liveness and data-flow. 
As we show in this chapter, this is not the case with structural versions.

In this chapter, we focus on three example applications. We look at structural liveness and
data-flow and we also consider calculus for bounded reuse, which checks how many times a variable
is accessed and generalizes linear logics (that restrict variables to be used exactly once).

\subsection{Contributions}

Compared to the previous chapter, the structural coeffect calculi we consider are more 
homogeneous and so finding the common pattern is in some ways easier. However, the systems
are somewhat more complicated as they need to keep annotations attached to individual variables. 
The contributions of this chapter are as follows:

\begin{itemize}
\item We present a \emph{structural coeffect calculus} with a type system that is parameterized by a 
  \emph{structural coeffect algebra} and can be instantiated to obtain all of the three examples
  discussed (Section~\ref{sec:struct-calculus}).

\item We give the equational theory of the calculus. We prove the type-preservation property 
  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}).

\item We show how to extend indexed comonads introduced in the previous section to 
  \emph{structural indexed comonads} and use them to give the semantics of structural 
  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
  categorical semantics provides a motivation for the design of the calculus.
\end{itemize}

@@

In the previous chapter, we discussed the correspondence between coeffects and effects (and 
between comonads and monads). As noted earlier, the $\lambda$-calculus is assymetric in that
an expression has multiple inputs (variables in the context), but just a single result 
(the resulting value) and so monads and effects have no notion directly corresponding to 
structural coeffect systems.

@@

Our work follows the language semantics style in that we provide a structural semantics to
the terms of ordinary $\lambda$-calculus. The most closely related work has been done in the
meta-langauge style, which extends the terms and types with constructs working with the context
explicitly. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
$A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
annotations as stand-alone types.

Structural coeffect systems annotate the whole variable context with a \emph{vector} of annotations.
For example, a context with variables $x$ and $y$ annotated with $\cclrd{s}$ and $\cclrd{t}$, 
respectively is written as $\coctx{x\!:\!\tau_1,y\!:\!\tau_2}{\alift{\cclrd{s},\cclrd{t}}}$.
This means that the typing judgements have the same structure as those of the flat coeffect
calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify the two 
systems and compose tracking of flat and structural properties.

@@

In the structural coeffect calculus, a vector of variables in the free-variable context is 
annotated with a vector of primitive (scalar) coeffect annotations. These annotations differ
for different coeffect calculi and their properties are captured by the \emph{structural 
coeffect scalar} definition below. The scalar annotations can be integers (how many past values we 
need) or annotations specifying whether a variable is live or not.

Scalar annotations are written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$ (following the style used in 
the previous chapter). Functions always have exactly one input variable and so they are annotated
with a coeffect scalar. Thus the expressions and types of structural coeffect calculi are the same
as in the previous chapter (except that annotation on function type is now a structural coeffect 
scalar):
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
In the previous chapter, the free variable context $\Gamma$ has been treated as a set. In the 
structural coeffect calculus, the order of variables matters. Thus we treat free variable context
as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
of the vector:
%
\begin{equation*}
\begin{array}{rcl}
 \Gamma &\narrow{=}& \langle x_1\!:\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
 \xlen{\Gamma} &\narrow{=}& n
\end{array}
\end{equation*}
%
For readability, we use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
typing judgements, but the free variable context should be understood as a vector.
Furthermore, the usual notation $\Gamma_1, \Gamma_2$ stands for the tensor product.
Given $\Gamma_1 = \langle x_1\!:\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
$\Gamma_2 = \langle x_{n+1}\!:\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.

The free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
we write the vectors of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging 
over vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face 
and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
of coeffects is provided by the \emph{structural coeffect algebra} definition below.

@@

The structural coeffect scalar structure is similar to \emph{flat coeffect algebra} with the 
exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
pointwise composition, as well as a relation $\cleq$ that defines sub-coeffecting.

@@

\noindent
In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
available from the declaration site and the call site or, in the syntactic reading, to split the 
context requirements.

In the structural coeffect calculus, we use a vector instead -- combining and splitting of 
coeffects becomes just vector a concatenation or splitting, respectively, which is provided
by the tensor product. The operations on vectors are indexed by integers representing the lengths
of the vectors. The additional structure required by the type system for structural coeffect 
calculi is given by the following definition.

@@

The type system for structural coeffect calculus is similar to sub-structural type systems
in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
weakening, exchange or contraction -- this is done by checking the types of sub-expressions
in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
allows weakening, exchange and contraction, but using explicit \emph{structural} rules that
perform corresponding transformation on the coeffect annoation.

\paragraph{Syntax-driven rules.}
The variable access rule (\emph{var}) annotates the corresponding variable as being used
using $\cunit$. Note that, as in sub-structural systems, the free variable context contains 
\emph{only} the accessed variable. Other variables can be introduced using explicit weakening.
Constants (\emph{const}) are type checked in an empty variable context, which is annotated with
an empty vector of coeffect annotations.

The (\emph{abs}) rule assumes that the free variable context of the body can be split into
a potentially empty \emph{declaration site} and a singleton context containing the bound
variable. The corresponding splitting is performed on the coeffect vector, uniquely associating
the annotation $\cclrd{s}$ with the bound variable $x$. This means that the typing rule removes
non-determinism present in flat coeffect systems.

@@

\paragraph{Structural rules.}
The remaining rules, shown in Figure~\ref{fig:struct-types} (b), are not syntax-directed.
They allow different transformation of the free variable context. We include sub-coeffecting
(\emph{sub}) as one of the rules, allowing sub-coeffecting on coeffect scalars belonging
to individual variables. The remaining rules capture \emph{weakening}, \emph{exchange} and
\emph{contraction} known from sub-structural systems.

The (\emph{weak}) allows adding a variable to the context, extending the coeffect
vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
in the context, performing the same reordering on the coeffect vector. Finally recall that 
variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
allows re-using a variable as we can type check sub-expressions using two separate varaibles and 
then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
is the only place in the structural coeffect system where context requiremens are combined,
or semantically, where the same context is shared.

@@

\caption{Type system for the structural coeffect calculus}
\label{fig:struct-types}
\end{figure}

@@

This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
as over- and under-approximations of $\atimes$. This is demonstrated by the (\emph{approximation})
requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
$\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$. Semantically, flat abstraction
combines available context, potentially discarding parts of it (under-approximation), while flat
application splits available context, potentially duplicating parts of it (over-approxi\-mation)\footnote{
Because of this duality, earlier version of coeffects in \cite{coeffects-icalp13}
used $\wedge$ and $\vee$.}.

Second, the operator $\cpar$ is used when the semantics passes the same context to multiple
sub-expressions. In flat systems, this happens in (\emph{app}) and (\emph{pair}), because the 
sub-expressions may share variables. In structural systems, this is separated into an 
explicit contraction rule.

\paragraph{Let binding.}
The other aspect that makes structural systems simpler is that they remove the need for 
separate let binding. As discussed in Section~\ref{sec:flat-calculus-let}, flat calculi include
let binding that gives a \emph{more precise} typing than combination of abstraction and 
application. This is not the case for structural coeffects.

\begin{remark}[Let binding]
\label{thm:structural-letbinding}
In a structural coeffect calculus, the typing of $(\lambda x.e_2)~e_1$ is equivalent
to the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$.
\end{remark}
\begin{proof}
Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that 
in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$. 
%
\begin{equation*}
\inference
  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
    \inference
      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
  { \inference
      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
\end{equation*}

@@

The structural coeffect calculus can be instantiated to obtain the structural coeffect calculi
presented in Section~\ref{sec:applications-structural}. Two of them -- structural data-flow and
structural liveness provide a more precise tracking of properties that can be tracked using 
flat systems. Formally, a flat coeffect algebra can be turned into a structural coeffect algebra
(by dropping the $\czip$ operator), but this does not always give us a meaningful system -- for
example, it is not clear why one would associate implicit parameters with individual variables.

On the other hand, some of the structural systems do not have a flat equivalent, typically 
because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
algebra. This is the case, for example, for the bounded variable use.

\begin{example}[Structural liveness]
The structural coeffect algebra for liveness is formed by
$(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is 
the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
with a join $\sqcup$ and meet $\sqcap$. 
\end{example}

\begin{example}[Structural data-flow]
In data-flow, context is annotated with natural numbers and the flat coeffect algebra is formed 
by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
\end{example}

\noindent
For the two examples that have both flat and structural version, obtaining the strucutral coeffect
algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The laws
required by a structural coeffect algebra are the same as those required by the flat version
and so the above definitions are both valid. Similar construction can be used for the
\emph{optimized data-flow} example from Section~\ref{sec:flat-calculus-examples}.

It is important to note that this gives us a systems with \emph{different} properties. The 
information are now tracked per-variable rather than for the enitre context. For data-flow,
we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
for a pointwise extension of the $+$ operator, such that
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} + k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
%
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
The rule appears similar to the flat one, but there is an important difference. Because of the
structural nature of the type system, it only increments the required number of values for 
variables that are used in the expression $e$. Annotations of other variables can be left 
unchanged.

@@

When introducing structural coeffect systems in Section~\ref{sec:applications-structural}, we 
included a concrete semantics of structural liveness and bounded variable reuse. In this section,
we generalize the examples using the notion of \emph{structural indexed comonad}, which is an 
extension of \emph{indexed comonad} structure. As in the previous chapter, the main aim of this 
section is to motivate and explain the design of the structural coeffect calculus shown in 
Section~\ref{sec:struct-calculus}. The semantics highlights the similarities and differences 
between the two systems.

Most of the differences between flat and structural systems arise from the fact that contexts
in structural coeffect systems are treated as \emph{vectors} rather than sets modelled using 
categorical products, so we start by discussing our treatment of vectors.

@@

In the flat coeffect calculus, the context is interpreted as a product and so a typing judgement
$\coctx{x_1\!:\tau_1, \ldots, x_n:\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a morphism 
$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model, we can 
freely transform the value contained in the context modelled using an indexed comonad
$\ctyp{\cclrd{r}}$. For example, the function $\ident{map}_{\cclrd{r}}~\pi_i$ transforms
a context $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a value 
$\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.

The ability to freely transform the variable structure is not desirable in the model of 
structural coeffect systems. Our aim is to guarantee (by construction) that the structure of
the coeffect annotations matches the structure of variables. To achieve this, we model vectors
using a structure distinct from ordinary products which we denote $-\xtimes-$. 
For example, the judgement $\coctx{x_1\!:\tau_1, \ldots, x_n:\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.

@@

\\[0.5em]

@@

  \hspace{2.5em}

@@

\\[1.5em]

@@

  \hspace{2em}

@@

\\[0.5em]

>>>>

In this chapter, we take the next step towards a unified theory of context-aware programming
languages. As already discussed, we attempt to find abstractions that are \emph{general enough} 
to capture a wide range of useful programming language features, but \emph{specific enough} to 
let us identify and capture interesting properties of the context.

In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware prorgamming
languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
calculus} in Chapter~\ref{ch:flat}. Here, we consider per-variable contextual properties
and we introduce the \emph{structural coeffect calculus}.

The flat coeffect system captures a number of interesting use-cases (implicit parameters, liveness 
and data-flow), but provides relatively imprecise approximation and weak syntactical properties. 
Some of those examples can be also seen as per-variable properties. For those, structural coeffect
systems give a more precise information about the context. However, we also look at other applications 
that arise from the work on sub-structural logics, a pathway to coeffects discussed in 
Section~\ref{sec:path-logic}.

As we will see, the categorical semantics of the structural coeffect calculus is based on the
same underlying structure as the semantics of the flat calculus. However, the syntactical properties
of the calculus are quite different and so it is worth considering it in a separate chapter.

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item[--] We present a \emph{structural coeffect calculus} as a type system that is parameterized 
  by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}. We show how the 
  system captures fine-grained liveness and data-flow information, as well as a calculus for bounded 
  reuse (checking how many times is a variable accessed).

\item[--] We extend indexed comonads introduced in the previous chapter to 
  \emph{structural indexed comonads} and use them to provide the semantics of structural 
  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
  semantics provides a motivation for the design of the calculus.

\item[--] We discuss the equational theory of the calculus. For structural coeffects, the substitution 
  lemma holds for all instances of the calculus. Thus, we prove the type-preservation property 
  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}). 
\end{itemize}

@@

Compared to the previous chapter, structural coeffect calculi we consider are more 
homogeneous and so finding the common pattern is in some ways easier. However, the systems
are more complicated as they need to keep annotations attached to individual variables
and thus require explicit structural rules. Before looking at the system, we briefly consider
the most important related work.

@@

In the previous chapter, we discussed the correspondence between coeffects and effects (and 
between comonads and monads). As noted earlier, the $\lambda$-calculus is asymmetric in that
an expression has multiple inputs (variables in the context), but just a single result 
(the resulting value) and so monads and effects have no notion directly corresponding to 
structural coeffect systems.

@@

Our work follows the ``language semantics'' style in that we provide a structural semantics to
the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in 
the meta-langauge style, which extends the terms and types of a language with constructs for explicitly
manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
$A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
annotations as stand-alone types (indexed comonads only appear in the semantics).

Our structural coeffect systems annotate the whole variable context with a \emph{vector} of annotations.
For example, a context with variables $x$ and $y$ annotated with $\cclrd{s}$ and $\cclrd{t}$, 
respectively is written as $\coctx{x\!:\!\tau_1,y\!:\!\tau_2}{\alift{\cclrd{s},\cclrd{t}}}$.
A benefit of this approach is that the typing judgements have the same structure as those of the 
flat coeffect calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify 
the two systems.

@@

In the structural coeffect calculus, a vector of variables forming the free-variable context is 
annotated with a vector of primitive (scalar) coeffect annotations. These annotations differ
for different coeffect calculi and their properties are captured by the definition of \emph{structural 
coeffect scalar} below. The scalar annotations can be \eg~integers (how many past values we 
need) or values of a two-point lattice specifying whether a variable is live or not.

Scalar annotations are written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$ (following the style used in 
the previous chapter). Functions always have exactly one input variable and so they are annotated
with a single coeffect scalar. Thus the expressions and types of structural coeffect calculi are the 
same as in the previous chapter. The only difference is that $\cclrd{r}, \cclrd{s}, \cclrd{t}$ now
range over values of a \emph{structural coeffect scalar} rather than over values of a \emph{flat 
coeffect algebra} (the definitions are related, but not the same).
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
In the previous chapter, the free variable context $\Gamma$ was treated as a set. In the 
structural coeffect calculus, the order of variables matters. Thus we treat a free variable 
context as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
of the vector:
%
\begin{equation*}
\begin{array}{rcl}
 \Gamma &\narrow{=}& \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
 \xlen{\Gamma} &\narrow{=}& n
\end{array}
\end{equation*}
%
For readability, we use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
typing judgements, but the free variable context should be understood as a vector.
Furthermore, the notation $\Gamma_1, \Gamma_2$ is here seen as the tensor product.
Given $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.

Free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
we write the vectors of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging 
over vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face 
and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
of coeffects is provided by the definition of \emph{structural coeffect algebra} below.

@@

The \emph{structural coeffect scalar} structure is similar to \emph{flat coeffect algebra} with the 
exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
pointwise composition, as well as a relation $\cleq$ that defines sub-coeffecting.

@@

\noindent
In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
available from the declaration site and the call site or, in the syntactic reading, to split the 
context requirements. The other difference is that the $\cpar$ operation of structural coeffect
scalar is not required to be idempotent. The bounded reuse example uses $\cpar$ which is not 
idempotent.

In the structural coeffect calculus, we use a vector instead -- combining and splitting of 
coeffects becomes just vector concatenation or splitting, respectively; these are provided
by the tensor product. The operations on vectors are indexed by integers representing the lengths
of the vectors. The additional structure required by the type system for structural coeffect 
calculi is given by the following definition.

@@

The type system for the structural coeffect calculus is similar to sub-structural type systems
in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
weakening, exchange or contraction -- this is done by checking the types of sub-expressions
in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
perform corresponding transformation on the coeffect annoation.

\paragraph{Syntax-driven rules.}
The syntax-driven rules of the type system are shown in Figure~\ref{fig:struct-types} (a).
The variable access rule (\emph{var}) annotates the corresponding variable as being accessed
using $\cunit$. Note that, as in sub-structural systems, the free variable context contains 
\emph{only} the accessed variable. Other variables can be introduced using explicit weakening.
Constants (\emph{const}) are type checked in an empty variable context, which is annotated with
an empty vector of coeffect annotations.

The (\emph{abs}) rule assumes that the free variable context of the body can be split into
a potentially empty \emph{declaration site} and a singleton context containing the bound
variable. The corresponding splitting is performed on the coeffect vector, uniquely associating
the annotation $\cclrd{s}$ with the bound variable $x$. This means that the typing rule removes
non-determinism of the type inference present in flat coeffect systems.

@@

\paragraph{Structural rules.}
The remaining rules, shown in Figure~\ref{fig:struct-types} (b), are not syntax-directed.
They allow different transformation of the free variable context. We include (\emph{sub}) 
as one of the rules, allowing sub-coeffecting on coeffect scalars belonging to individual 
variables. The remaining rules capture \emph{weakening}, \emph{exchange} and \emph{contraction} 
known from sub-structural systems.

Rule (\emph{weak}) allows adding a variable to the context, extending the coeffect
vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
in the context, performing the same reordering on the coeffect vector. Finally recall that 
variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
allows re-using a variable. We can type check sub-expressions using two separate varaibles and 
then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
is the only place in the structural coeffect system where context requiremens are combined
(semantically, this is where the available context is shared).

@@

\figcaption{Type system for the structural coeffect calculus}
\label{fig:struct-types}
\end{figure}

@@

This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
as over-approximations and under-approximations of $\atimes$. This is demonstrated by the 
(\emph{approximation}) requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
$\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$. Semantically, flat abstraction
combines two values representing the available context, potentially discarding parts of it 
(under-approximation), while flat application splits the available context (a single value), 
potentially duplicating parts of it (over-approxi\-mation)\footnote{Because of this duality, earlier 
version of coeffect systems \cite{coeffects-icalp13} used $\wedge$ and $\vee$.}.

Secondly, the operator $\cpar$ is used when the semantics passes a given context to multiple
sub-expressions. In flat systems, the context is shared in (\emph{app}) and (\emph{pair}), because 
the sub-expressions may share variables. In structural systems, the sharing is isolated into an 
explicit contraction rule.

\paragraph{Let binding.}
The other aspect that makes structural systems simpler is that they remove the need for 
separate let binding. As discussed in Section~\ref{sec:flat-calculus-let}, flat calculi include
let binding that gives a \emph{more precise} typing than combination of abstraction and 
application. Structural coeffect systems avoid this issue.

\begin{remark}[Let binding]
\label{thm:structural-letbinding}
In a structural coeffect calculus, the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$ can be
seen as a derived rule, \ie~its typing is equivalent to the typing of the expression
$(\lambda x.e_2)~e_1$.

\end{remark}
\begin{proof}
Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that 
in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$. 
%
\begin{equation*}
\inference
  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
    \inference
      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
  { \inference
      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
\end{equation*}

@@

The structural coeffect calculus above can be instantiated to obtain the 3 structural coeffect calculi
presented in Section~\ref{sec:applications-structural}. Two of them -- structural data-flow and
structural liveness provide a more precise tracking of properties that can be tracked using 
flat systems. Formally, any flat coeffect algebra can be turned into a structural coeffect scalar
(by dropping the $\czip$ operator), but this does not always give us a meaningful system -- for
example, implicit parameters could be attached to individual variables, but, to our knowledge, this
has not been used in any practical system.

On the other hand, some of the structural systems do not have a flat equivalent, typically 
because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
scalar. This is the case, for example, for the bounded variable use.

\begin{example}[Structural liveness]
The structural coeffect scalar for liveness is formed by
$(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is 
the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
with a join $\sqcup$ and a meet $\sqcap$. 
\end{example}

\begin{example}[Structural data-flow]
In data-flow, context is annotated with natural numbers and the structural coeffect scalar is formed 
by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
\end{example}

\noindent
For the two examples that have both flat and structural version, obtaining the structural coeffect
algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The laws
required by a structural coeffect algebra are the same as those required by the flat version
and so the above definitions are both valid. Similar construction can be used for the
\emph{optimized data-flow} example from Section~\ref{sec:flat-calculus-examples}.

It is important to note that this gives us a systems with \emph{different} properties. The 
information are now tracked per-variable rather than for enitre contexts. For data-flow,
we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
for a pointwise extension of the $+$ operator, such that
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} \aclrd{+} k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
%
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
The rule appears similar to the flat one, but there is an important difference. Because of the
structural nature of the type system, it only increments the required number of values for 
variables that are used in the expression $e$. Annotations of other variables can be left 
unchanged.

@@

When we introduced structural coeffect systems in Section~\ref{sec:applications-structural}, we 
included a concrete semantics of structural liveness and bounded variable reuse. In this section,
we generalize the examples using the notion of \emph{structural indexed comonad}, which is an 
extension of the \emph{indexed comonad} structure. Similarly to \emph{flat indexed comonad}, the
structural variant adds operations that are needed to embed full $\lambda$-calculus, this time 
with per-variable contexts.

As in the previous chapter, the main aim of this section is to motivate and explain the design of 
the structural coeffect calculus shown in Section~\ref{sec:struct-calculus}. However, the semantics 
also highlights the similarities and differences between the flat and structural system.
Most of the differences arise from the fact that contexts in structural coeffect systems are treated 
as \emph{vectors} rather than sets.

@@

Recall that in the flat coeffect calculus, the context is interpreted as a product and so a typing 
judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a 
morphism $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model, 
we can freely transform the value contained in the context modelled using an indexed comonad
$\ctyp{\cclrd{r}}$.

Previously, we defined the $\ident{map}$ function (in terms of $\ident{cobind}$ and $\ident{counit}$),
which transforms the value inside the context without affecting the coeffect annotation. Thus,
we can use $\ident{map}_{\cclrd{r}}~\pi_i$ to transform a context containing product of variables
$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a context containing a single value 
$\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.

The ability to freely transform the variable structure is not desirable in the model of 
structural coeffect systems. Our aim is to guarantee (by construction) that the structure of
the coeffect annotations matches the structure of variables. To achieve this, we model vectors
using a structure distinct from ordinary products which we denote $-\xtimes-$. 
For example, the judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.
We assume that the operator is equipped with necessary associativity transformations allowing us
to use it freely on more than two values.

@@

\\[0.4em]

@@

  \hspace{1.5em}

@@

\\[1.4em]

@@

  \hspace{1.75em}

@@

\\[0.0em]
